Synthesizing Quotient Lenses

### 2017-10-18 papers/QuotientLensSynth/main.tex
some work on the intro to move the code into figures

\usepackage{amsmath, amssymb, verbatim, enumerate, graphicx, centernot, tikz,
array, mathtools, bussproofs, stmaryrd, enumitem, stackengine}

@@

The code sample below shows how the programmer might specify these equivalence
relations on the \bibtex{} records in Boomerang using canonizers:\bcp{Too
  big to display in-line.  Better to put it in a figure.}

\begin{verbatim}
let project (r : regexp) (s : string) = canonizer_of_lens ((del r) . (ins s))

let wsp : regexp = [ \n\t\r]+

let wsp_to_space : canonizer = project wsp " "

let valid_name : canonizer = 
name | (name . wsp_to_space . (name .wsp_to_space)* .name)

let wsp_to_and : canonizer = project (wsp . "and" . wsp) " and "

let bib_names : canonizer =

valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)* . valid_name)

let bib_author : canonizer = "author = \"" . bib_names . "\""

let label : regexp = [a-zA-Z0-9]+

let title : canonizer = 
word | (word . wsp_to_space . (word . wsp_to_space)* . word)

let bib_title : canonizer = "title = {" . title . "}"

let wsp_to_comma : canonizer = project ("," . wsp) ",\n"

let first_perm : canonizer = bib_author . wsp_to_comma . bib_title

let second_perm : canonizer = bib_title . wsp_to_comma . bib_author

let bibtex : canonizer = 
"@book{" . label. wsp_to_comma . (first_perm | second_perm) . "}"
\end{verbatim}

Using the in-built \verb!canonizer_of_lens! function which extracts a canonizer
from a lens, it is easy to define a an equivalence relation that sends all
members of a regular language to a canonical representative. For example
the expression 
\begin{verbatim}
let wsp_to_space : canonizer = canonizer_of_lens ((del wsp) . (ins " "))
\end{verbatim}

\noindent defines a canonizer which sends arbitrary occurences of whitespace
characters to a single space. However, a problem arises when trying to specify
the equivalence relation which nests this projection relation with the relation
that considers all permutations of the \bibtex{} fields equivalent.

@@

\begin{verbatim}
let first_perm : canonizer = bib_author . wsp_to_comma . bib_title

let second_perm : canonizer = bib_title . wsp_to_comma . bib_author
\end{verbatim}

One possible solution is to permute the alternative canonizers

\noindent \verb!bib_author. wsp_to_comma! and \verb!bib_title . wsp_to_comma!,
but Boomerang's type-checker will reject this expression since the permutation
combinator accepts regular expressions rather than canonizers. In any case, the
regular expression
\begin{verbatim}
bib_author . wsp_to_comma . bib_title . wsp_to_comma | 
bib_title . wsp_to_comma . bib_author . wsp_to_comma
\end{verbatim}
is not equivalent to,
\begin{verbatim}
bib_author . wsp_to_comma . bib_title  | 
bib_title . wsp_to_comma . bib_author
\end{verbatim}
thus the programmer would have to write more lenses and canonizers to correct
this. Therefore, in order to utilize this trick, the programmer might choose to
specify all the regular expressions first (including the whole regular
expressions canonized by the \verb!bib_author! and \verb!bib_title!
canonizers), and then the canonizers separately. Alternatively, he/she may stick
to using canonizers and then liberally make use of the built-in
functions \verb!uncanonized_type! and \verb!canonized_type! which return the
regular expression matching all the data and the regular expression matching
the canonical representatives respectively.

@@

With QREs we can specify \bibtex{} records as follows:\bcp{Again, belongs in
a figure.  And, since both figures are long, maybe there is a visual way to
highlight the differences...?}

\begin{verbatim}
let wsp_to_space : canonizer = project [ \n\t\r]+ -> " "

let valid_name : canonizer = 
name | (name . wsp_to_space . (name . wsp_to_space)* . name)

let wsp_to_and : canonizer = 
project ([ \n\t\r]+ . "and" . [ \n\t\r]+) -> " and "

let bib_names : canonizer =

valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)* . valid_name)

let bib_author : canonizer = "author = \"" . bib_names . "\""

let label : regexp = [a-zA-Z0-9]+

let title : canonizer = word | (word . wsp_to_space . (word . wsp_to_space)* .
word)

let bib_title : canonizer = "title = {" . title . "}"

let bib_perm  : canonizer = perm (label, bib_author, bib_title) with
(project ("," . [ \n\t\r]* ) -> ",\n")

let bibtex : canonizer = "@book{" . bib_perm . "}"
\end{verbatim}

The expression \verb!project [ \n\t\r]+ -> " "! is a QRE which matches any
number of whitespace characters, and then considers all strings of whitespace
characters as equivalent with a single space as the canonical representative. 
Another QRE in the example above is the

\begin{center}
\begin{verbatim}
perm (label, bib_author, bib_title) with (project ("," . [\n\t\r]* ) -> ",\n")
\end{verbatim}
\end{center}

\noindent QRE, which specifies that the label, title and author fields can occur
in any order, with the fields separated by a comma followed by arbitrary amounts
of whitespace. The canonical representation is that in which the label, author
and title fields occur in that order, separated by a comma followed a newline.
The \verb!perm! QRE is defined at the level of canonizers which allows for
easy nesting of permutations. Similarly, it is easy to specify any number of
fields within the \verb!perm! QRE.

While QREs kill two birds with one stone by specifying the
data and the equivalence relation on the data in one expression, he/she
still has to manually program the underlying lens which maps the equivalence
classes of the source data to the equivalence classes of the view data, a task
which is itself tedious and error-prone as the code snippet below illustrates:
\begin{verbatim}
let preamble : (lens in "@book{" <=> "%0 Book\n%F ") = 
    del "@book{" . ins "%0 Book\n%F "

let author_lens : (lens in (",\n" . bib_author) <=> ("\n" . end_author)) = 
    del ",\nauthor = \""
    . ins "\n%A "
    . valid_name
    . (del " and "
         . ins "\n"
         . (ins "%A " . valid_name . del " and " . ins "\n")*  . ins "%A "
         . valid_name
           || "")

let title_lens : 
(lens in ("\",\n" . bib_title . "},\n}") <=> ("\n" . end_title)) =
del "\",\ntitle = {" . ins "\n%T " . title . del "},\n}"

let bib_to_end : (lens in bib_tex <=> end_note) =
preamble . label . author_lens . title_lens
\end{verbatim}

This lens transforms the \bibtex{} data to the EndNote (and vice versa) data by
matching the delimiters and separators in a bijective manner. For example,
the \verb!preamble! lens replaces the \bibtex{} preamble \verb!@book! with the
EndNote preamble \verb!%0 Book\n%F ! in the \bibtex{record}. The programmer
can ill afford to make a mistake with any one of the small but vital details
of the transformations since even a one character mistake might render the lens
incorrect. This makes lens programming tedious and error-prone.

@@

We have already encountered the \verb!project R -> s! QRE which sends all
elements of the language described by the regular expression $R$, and the

\noindent \verb!perm (q_1, ..., q_n) with q! QRE which sends all permutations of
the QREs $q_1, \ldots, q_n$ separated by the QRE $q$ to the canonical
QRE $q_1 \cdot q \cdot \ldots \cdot q \cdot q_n$.

@@

\begin{verbatim}
let squash_author : canonizer =  
squash space_name -> comma_name using (get comma_to_space)

@@

\begin{verbatim}
let ambiguous : canonizer = "a"* . (project "a"* -> "a")
\end{verbatim}

>>>>

\usepackage{amsmath, amssymb, verbatim, enumerate, graphicx, centernot, tikz,
array, mathtools, bussproofs, stmaryrd, enumitem, stackengine}
\usepackage{relsize}
\usepackage{listings}
\lstset{
  language=Caml,
  basicstyle=\upshape\sffamily,
  keywordstyle=\upshape\sffamily\color{dkblue},
  keepspaces=true,
  framexleftmargin=1ex,
  framexrightmargin=1ex,
  showstringspaces=false,
  commentstyle=\itshape\rmfamily,
  columns=flexible,
  stringstyle=\sffamily\color{dkred},
  mathescape
}
\newcommand{\cd}[1]{\lstinline[backgroundcolor=\color{white}]$#1$}

@@

The code sample in Figure~\ref{fig:example-canonizers} shows how the programmer
might specify these equivalence relations on the \bibtex{} records in Boomerang
using canonizers.

\begin{figure}[t]

\begin{lstlisting}
let project (r : regexp) (s : string) = canonizer_of_lens ((del r) . (ins s))
let wsp : regexp = [ \n\t\r]$^+$
let wsp_to_space : canonizer = project wsp " "
let valid_name : canonizer = 
    name | (name . wsp_to_space . (name .wsp_to_space)$^*$ .name)
let wsp_to_and : canonizer = project (wsp . "and" . wsp) " and "
let bib_names : canonizer =  
    valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)$^*$ . valid_name)
let bib_author : canonizer = "author = \"" . bib_names . "\""
let label : regexp = [a-zA-Z0-9]$^+$
let title : canonizer = 
    word | (word . wsp_to_space . (word . wsp_to_space)$^*$ . word)
let bib_title : canonizer = "title = {" . title . "}"
let wsp_to_comma : canonizer = project ("," . wsp) ",\n"
let first_perm : canonizer = bib_author . wsp_to_comma . bib_title
let second_perm : canonizer = bib_title . wsp_to_comma . bib_author
let bibtex : canonizer = 
    "@book{" . label. wsp_to_comma . (first_perm | second_perm) . "}"
\end{lstlisting}

  \caption{Boomerang canonizers for the \bibtex{}--EndnNote equivalence
    relations.}
  \label{fig:example-canonizers}
\end{figure}

Using the in-built \cd{canonizer_of_lens} function which extracts a canonizer
from a lens, it is easy to define an equivalence relation that sends all
members of a regular language to a canonical representative. For example
the expression 
\begin{lstlisting}
let wsp_to_space : canonizer = canonizer_of_lens ((del wsp) . (ins " "))
\end{lstlisting}
\noindent defines a canonizer which sends arbitrary occurences of whitespace
characters to a single space. However, a problem arises when trying to specify
the equivalence relation which nests this projection relation with the relation
that considers all permutations of the \bibtex{} fields equivalent.

@@

\begin{lstlisting}
let first_perm : canonizer = bib_author . wsp_to_comma . bib_title
let second_perm : canonizer = bib_title . wsp_to_comma . bib_author
\end{lstlisting}

One possible solution is to permute the alternative canonizers
\cd{bib_author.wsp_to_comma} and \cd{bib_title.wsp_to_comma},
but Boomerang's type-checker will reject this expression since the permutation
combinator accepts regular expressions rather than canonizers. In any case, the
regular expression
\begin{lstlisting}
bib_author . wsp_to_comma . bib_title . wsp_to_comma | 
bib_title . wsp_to_comma . bib_author . wsp_to_comma
\end{lstlisting}
is not equivalent to,
\begin{lstlisting}
bib_author . wsp_to_comma . bib_title  | 
bib_title . wsp_to_comma . bib_author
\end{lstlisting}
thus the programmer would have to write more lenses and canonizers to correct
this. Therefore, in order to utilize this trick, the programmer might choose to
specify all the regular expressions first (including the whole regular
expressions canonized by the \cd{bib_author} and \cd{bib_title}
canonizers), and then the canonizers separately. Alternatively, he/she may stick
to using canonizers and then liberally make use of the built-in
functions \cd{uncanonized_type} and \cd{canonized_type} which return the
regular expression matching all the data and the regular expression matching
the canonical representatives respectively.

@@

With QREs we can specify \bibtex{} records as shown in Figure~\ref{fig:example-qre}.

\begin{figure}[t]
\begin{lstlisting}
let wsp_to_space : canonizer = project [ \n\t\r]$^+$ $\to$ " "
let valid_name : canonizer = 
   name | (name . wsp_to_space . (name . wsp_to_space)$^*$ . name)
let wsp_to_and : canonizer = 
   project ([ \n\t\r]$^+$ . "and" . [ \n\t\r]$^+$) $\to$ " and "
let bib_names : canonizer =  
   valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)$^*$ . valid_name)
let bib_author : canonizer = "author = \"" . bib_names . "\""
let label : regexp = [a-zA-Z0-9]$^+$
let title : canonizer = word | (word . wsp_to_space . (word . wsp_to_space)$^*$ . word)
let bib_title : canonizer = "title = {" . title . "}"
let bib_perm  : canonizer = perm (label, bib_author, bib_title) with
   (project ("," . [ \n\t\r]$^*$ ) $\to$ ",\n")
let bibtex : canonizer = "@book{" . bib_perm . "}"
\end{lstlisting}

  \caption{QRE specification of \bibtex{} records.}
  \label{fig:example-qre}
\end{figure}

The expression \cd{project [ \n\t\r]+ -> " "} is a QRE that matches any
number of whitespace characters, and then considers all strings of whitespace
characters as equivalent with a single space as the canonical representative. 
Another QRE in the example above is the

\begin{center}
\begin{lstlisting}
perm (label, bib_author, bib_title) with (project ("," . [\n\t\r]$^*$ ) $\to$ ",\n")
\end{lstlisting}
\end{center}

\noindent QRE, which specifies that the label, title and author fields can occur
in any order, with the fields separated by a comma followed by arbitrary amounts
of whitespace. The canonical representation is that in which the label, author
and title fields occur in that order, separated by a comma followed a newline.
The \cd{perm} QRE is defined at the level of canonizers which allows for
easy nesting of permutations. Similarly, it is easy to specify any number of
fields within the \cd{perm} QRE.

While QREs kill two birds with one stone by specifying the
data and the equivalence relation on the data in one expression, he/she
still has to manually program the underlying lens which maps the equivalence
classes of the source data to the equivalence classes of the view data, a task
which is itself tedious and error-prone as the code snippet below illustrates:

\begin{lstlisting}
let preamble : (lens in "@book{" $\Leftrightarrow$ "%0 Book\n%F ") = 
    del "@book{" . ins "%0 Book\n%F "

let author_lens : (lens in (",\n" . bib_author) $\Leftrightarrow$ ("\n" . end_author)) = 
    del ",\nauthor = \""
    . ins "\n%A "
    . valid_name
    . (del " and "
         . ins "\n"
         . (ins "%A " . valid_name . del " and " . ins "\n")$^*$ . ins "%A "
         . valid_name
           || "")

let title_lens : (lens in ("\",\n" . bib_title . "},\n}") $\Leftrightarrow$ ("\n" . end_title)) =
   del "\",\ntitle = {" . ins "\n%T " . title . del "},\n}"

let bib_to_end : (lens in bib_tex $\Leftrightarrow$ end_note) =
   preamble . label . author_lens . title_lens
\end{lstlisting}

This lens transforms the \bibtex{} data to the EndNote (and vice versa) data by
matching the delimiters and separators in a bijective manner. For example,
the \cd{preamble} lens replaces the \bibtex{} preamble \cd{@book} with the
EndNote preamble \cd{\%0 Book\n\%F } in the \bibtex{record}. The programmer
can ill afford to make a mistake with any one of the small but vital details
of the transformations since even a one character mistake might render the lens
incorrect. This makes lens programming tedious and error-prone.

@@

We have already encountered the \verb!project R $\to$ s! QRE which sends all
elements of the language described by the regular expression $R$, and the

\noindent \cd{perm (q_1, ..., q_n) with q} QRE which sends all permutations of
the QREs $q_1, \ldots, q_n$ separated by the QRE $q$ to the canonical
QRE $q_1 \cdot q \cdot \ldots \cdot q \cdot q_n$.

@@

\begin{verbatim}
let squash_author : canonizer =  
squash space_name $\to$ comma_name using (get comma_to_space)

@@

\begin{verbatim}
let ambiguous : canonizer = "a"* . (project "a"* $\to$ "a")
\end{verbatim}
